---
title: "imputeR Documentation"
author: "Jinliang Yang"
date: "October 8, 2015"
output:
  pdf_document:
    number_sections: yes
    toc: yes
  html_document:
    toc: yes
  word_document: default
---

\newpage

# Introduction

## Crossing Scheme and Experimental Design

![alt text](scheme.pdf)

In this experiment, we selfed and outcrossed a set of ~70 teosinte landraces to get a progeny array composed of 4,875 individuals. The ~70 founders and all the progeny were genotyped using GBS. We also re-sequenced 20/70 founder lines (the others will be re-sequenced soon). Because of the high error rate of the GBS data, especially problematic for calling heterozygous sites, we employed a phasing and imputation strategy to infer the expected genotypes by combining parentage and GBS information.

This file is to document the `R` package we developed to solve the problems step by step.

## Install and Usage

Install devtools first, and then use devtools to install imputeR from github.

```{r, eval=FALSE}
# install and load devtools
devtools::install_github("hadley/devtools")
library(devtools)
# install and load imputeR
install_github("yangjl/imputeR")
library(imputeR)
```

## How to find help.
Within "R" console, type `?impute_parent` or `help(impute_parent)` to find help information about the function.

```{r}
?impute_parent
```





\newpage

# Infer parent's genotype from GBS data

If we have parent's WGS data, this step can be skipped.

We have observed mom and observed (selfed) kids.  We want to know $P(G|\theta)$, or the probability of  mom's genotype given observed data $\theta$. And according to [Bayes' theorem](https://en.wikipedia.org/wiki/Bayes%27_theorem), 

$P(G|\theta) \propto P(\theta|G) \times P(G)$,   

where $P(G)$ is the probability of the genotype according to the Hardy-Weinberg equilibrium estimated from the population.
This consists of observed genotypes ($G'$) of both mom and kids. So:

$P(G|\theta)\propto \left( \prod\limits_{i=1}^{k}{P(G'_k|G)} \right) \times P(G'_{mom}|G) \times P(G)$,  

where $P(G'_{mom}|G)$ is the probability of mom's observed genotype given a true genotype $G$ by considering error rates, i.e. GBS homozygote error = 0.02 and heterozygote error = 0.8.  
And $P(G'_{k}|G)$ is the probability of the $k$ th kid’s observed genotype given genotype $G$ by considering error rates and Mendelian segregation rate. The function `impute_mom` was implemented to compute mom's genotype probabilities.

## Example: step by step
In the below toy example, we simulate a family of self and outcross progeny with 50 kids for 3 loci. We will assume 50% missing data, and 50% selfing rate. 

Load the packages.
```{r, eval=TRUE}
library(devtools)
library(imputeR)
```

First we set some parameters for simulation
```{r}
    source("~/Documents/Github/imputeR/R/utils.R")
    source("~/Documents/Github/imputeR/R/ImputeParent.R")
    misscode = 3 # code for missing data
    numloci=3 # number of loci
    hom.error=0.02 #homozygous error rate
    het.error=0.8 #heterozygous error rate
    imiss=0.5 # % missing data
    selfing=0.5 # selfing rate
    size.array=50 # family size
    rec=0.25 # mean number of crossovers per chromosome
```

Then we make our focal parent
```{r}

    sfs <- getsfs() # make neutral SFS
    p <- sample(sfs, numloci, replace=TRUE) # get sample of allele freqs from SFS
    
    ### make focal_parent using a data.frame
    sim_focal <- data.frame(hap1=ran.hap(numloci,p), hap2=ran.hap(numloci,p))
```


What is our focal parent's real genotype:
```{r}
    sim_focal
```


Now we make our set of parents for each of our progeny. The first `outcrossed` parents are random, then next `size.array - outcrossed` are just the focal parent.
```{r}
    #first outcrossed
    outcrossed=rbinom(n=1,prob=(1-selfing),size=size.array)
    out_parents <- vector("list", outcrossed)
    out_parents <- lapply(1:outcrossed, function(i) data.frame(hap1=ran.hap(numloci,p), hap2=ran.hap(numloci,p)) )
    #now selfed
    self_parents <- vector("list", size.array-outcrossed)
    self_parents <- lapply(1:(size.array-outcrossed), function(i) sim_focal )
    #combine
    if(outcrossed==0){
        parent_array=self_parents  
    }else if (outcrossed==size.array){
        parent_array=out_parents
    }else{
        parent_array=c(out_parents,self_parents)
    }

    #now we make their diploid genotypes, we add the focal parent on to the end of the parents array
    parents<-lapply(parent_array, function(q) q[,1]+q[,2]  )
    parents[[size.array+1]]=c(sim_focal[,1]+sim_focal[,2])
    #finally, add error to make some crappy gbs_parents
    gbs_parents=lapply(parents, function(a) add_error(a,hom.error,het.error))
```

Now we make a progeny array for these parents.
```{r}
    progeny <- vector("list", size.array)
    #use the kid function!
    #each entry in progeny list has two vectors. [[1]] is true genotype, [[2]] is observed
    progeny <- lapply(1:size.array, function(a) kid(p2=list(parent_array[[a]][,1], parent_array[[a]][,2]), p1=list(sim_focal[,1],sim_focal[,2]), het.error, hom.error, rec, imiss, misscode))
    
    #now setup observed kids
    obs_kids=list()
    for(i in 1:size.array){ obs_kids[[i]]=progeny[[i]][[2]] }
```

Now we impute the focal parent
```{r}
    #which parent is our focal one? Here we set to end of parents array for ease
    obs_parent=size.array+1 #focal parent
    #which parents are the other parent of each offspring. These are in order since we simulated them that way.
    other_parents=c(1:outcrossed,rep(obs_parent,size.array-outcrossed)) #list of other parents

```


## Simulation using `imputeR` package

Above simulation steps were packed into a funcion `sim.array`. 
```{r}
# find help
?sim.array
test <- sim.array(size.array=50, numloci=100)
```

Now impute parent genotypes using `impute_parent` and extract results using `parentgeno`. In the resulting table `res`, the first three columns are the probabilities of genotype `0, 1, 2`. The 4th column is the odd ratio of the highest divided by the 2nd highest probability. `gmax` parent's genotype with the highest probability. `gor` parent's genotype with the highest probability and `OR` bigger than the specified threshold.

```{r, eval=FALSE}
#The stuff:
inferred_geno_likes <- impute_parent(test, hom.error=0.02, het.error=0.8, imiss=0.5)
res <- parentgeno(inferred_geno_likes, oddratio=0.6931472, returnall=TRUE)
res$true_parent <- test@true_parents[[50]]$hap1 + test@true_parents[[50]]$hap2
#error rates
nrow(subset(res, gmax != true_parent ))/nrow(res)
nrow(subset(res, gor != true_parent & gor !=3 ))/nrow(res)
```





## Real Data


To load `hdf5` file, you need to install Vince's [tasselr](https://github.com/vsbuffalo/tasselr) and [ProgenyArray](https://github.com/vsbuffalo/ProgenyArray) packages. If you fail to install them, please follow the above links and install the required dependencies. 


```{r, eval=FALSE}
# install devtools and then install the devlopmental version 
# of tasselr and ProgenyArray using devtools. 
devtools::install_github("hadley/devtools")
library(devtools) 
install_github("vsbuffalo/tasselr") 
install_github("vsbuffalo/ProgenyArray")
install_github("yangjl/imputeR")
```

Load the required packages. Note you have to specify the locations of the packages if they were not in your searching path.
```{r, eval=FALSE}
# load packages
library(parallel)
library(devtools)
options(mc.cores=NULL)
# you need to specify the location where the packages were installed. 
load_all("~/bin/tasselr")
load_all("~/bin/ProgenyArray")
load_all("~/Documents/Github/imputeR")
```

The following several lines help you to reformat the `*.h5` HDF5 file into an R object. You need to specify the path of the HDF5 file.

```{r, eval=FALSE}
# Note: at least 100G memory will be needed to load the hdf5 file
# load h5file
teo <- initTasselHDF5("largedata/teo.h5", version="5")
teo <- loadBiallelicGenotypes(teo, verbose = TRUE)
# reformat to imputeR object
source("R/load_data.R")
ob <- imputeRob(teo)
save(file="largedata/teo.RData", list="ob")  
```


Then, for each parent, run `impute_parent` as in the toy example above. Note that you will need to supply a vector of allele frequencies at each locus estimated from the parents. If you do not supply this or leave `p=NULL`, a random allele frequency drawn from the neutral SFS will be used instead (this is Very Bad).  Since parents are coded as 0,1, or 2 for $N$ parents the allele frequency $p$ at a locus can be calculated as $\frac{\sum_{i=1}^Np_i}{2N}$.




\newpage

# Phasing focal parent's haplotype 

According to Bayes' theorem, we got:

$$P(H_f|\theta) \propto P(\theta|H_f) \times P(H_f)$$  

- Where $\theta$ denotes observed data.
- $P(H_f)$ is the probability of our focal parent's haplotypes for a given window size of $n$.

Our prior assumption for $P(H_f)$ is that all possible haplotypes are equally likely. So,

$$P(H_f | \theta) \propto P(\theta | H_f)$$  

We assume $H_i$ as the haplotype of the other parent for the $i$th kid. Because all possible haplotypes ($H_i$, p ranged from 1 to k) inherited from other parents constitute a partition of the sample space, which includes $k$ kids reproduced by the focal parent ($H_f$). The hapoltypes ($H_i$) are pairwise mutually exclusive. Therefore,

$$P(H_f|\theta) \propto \sum_{i=1}^{k} P(\theta | H_f, H_i)$$

It can be further expressed by:
$$P(H_f|\theta) \propto \sum_{i=1}^{k} P(K'_i | H_f, H_i)$$ 

$$P(H_f|\theta) \propto  \sum_{i=1}^{k}\prod\limits_{l=1}^{n}{P(G'_{i,l}|H_1, H_2)} $$


- $P(K'_i|H_f, H_i)$ is the probability of $i$th kid's haplotype given parents' haplotype of $H_f$ and $H_i$.
- $P(G'_{i,l} | H_f, H_i)$ is the probability of $i$th kid's genotype at locus $l$ given parents' haplotype of $H_f$ and $H_i$.

We implemented a function `PhaseParent` to carry out the phasing work.

```{r}
maxlog_hap_ockid <- function(dad_hap, mom_geno, kid_geno){
    
    ### all possible mom haplotypes based on mom's genotype
    het_idx <- which(mom_geno==1)
    if(length(het_idx) > 0){
        haps1 <- setup_haps(win_length=length(het_idx))
        haps2 <- lapply(1:length(haps1), function(x) 1-haps1[[x]])
        allhaps <- c(haps1, haps2)
        mom_haps <- vector("list", 2^length(het_idx))
        mom_haps <- lapply(1:length(mom_haps), function(x) {
            temhap <- mom_geno/2
            temhap[het_idx] <- allhaps[[x]]
            return(temhap)})
    }else{
        mom_haps <- list(mom_geno/2)
    }   
    
    ### all possible genotypes of a kid
    allgeno1 <- lapply(1:length(mom_haps), function(x) dad_hap + mom_haps[[x]])
    allgeno2 <- lapply(1:length(mom_haps), function(x) (1-dad_hap) + mom_haps[[x]])
    allgenos <- c(allgeno1, allgeno2)
    
    ### return the possibility of the max genotype
    geno_probs <- lapply(1:length(allgenos), function(geno){
        sum( unlist(lapply(1:length(dad_hap), function(zz) {
            #log(probs[[2]] is the log prob. of kid's obs geno 
            #given the current phased geno and given dad is het. (which is why probs[[2]])
            tem <- probs[[2]][[ mom_geno[zz]+1 ]][allgenos[[geno]][zz]+1, kid_geno[zz]+1]
            return(log(tem))
        })))   
    })
    return(max(unlist(geno_probs)))
}

```

## Simulation

Above simulation steps were packed into a funcion `sim.array`. 
```{r}
test <- sim.array(50, 100, hom.error = 0, het.error = 0, rec = 0.25, selfing = 0.5, imiss = 0.1, misscode = 3)






```

Now impute parent genotypes using `impute_parent` and extract results using `parentgeno`. In the resulting table `res`, the first three columns are the probabilities of genotype `0, 1, 2`. The 4th column is the odd ratio of the highest divided by the 2nd highest probability. `gmax` parent's genotype with the highest probability. `gor` parent's genotype with the highest probability and `OR` bigger than the specified threshold.

```{r}
#The stuff:
inferred_geno_likes <- impute_parent(test, hom.error=0.02, het.error=0.8)
res <- parentgeno(inferred_geno_likes, oddratio=0.6931472, returnall=TRUE)
res$true_parent <- test@true_parents[[50]]$hap1 + test@true_parents[[50]]$hap2
#error rates
nrow(subset(res, gmax != true_parent ))/nrow(res)
nrow(subset(res, gor != true_parent & gor !=3 ))/nrow(res)
```





## Real Data




----------

# Imputing and Phasing Kids  
$P(H_k|\theta) \propto P(\theta|H_k) \times P(H_k)$   
$P(H_k|\theta) \propto \left( \prod\limits_{i=k}{P(H'_k|H_k)} \right) \times P(H_k)$  
$P(H_k|\theta) \propto \left( \prod\limits_{i=k}\prod\limits_{l=1}^{n}{P(G'_{i,l}|H_k)} \right) \times P(H_k)$

- Where $\theta$ denotes observed data.
- $P(H)$ is the probability of the haplotype for a given window size of $n$.
- $P(G'_{i,l}|H)$ is the probability of kid $i$ at locus $l$ for a given haplotype $H$.
- The prior $P(H)$ is that all possible haplotypes of a given window size are equally likely.

```{r, eval=FALSE}
phase <- read.csv("../data/sim_phasing_res.csv")

hist(phase$er, breaks=30, main="Simulation (N=100)",col="#faebd7", xlab="Phasing Error Rate")
abline(v=mean(phase$er), col="red", lwd=2)
abline(v=median(phase$er), col="darkblue", lwd=2)
```



----------
