---
title: "imputeR Documentation"
author: "Jinliang Yang"
date: "October 8, 2015"
output:
  pdf_document:
    number_sections: yes
    toc: yes
  html_document:
    toc: yes
  word_document: default
---

\newpage

# Introduction

## Crossing Scheme and Experimental Design

![alt text](scheme.pdf)

In this experiment, we selfed and outcrossed a set of ~70 teosinte landraces to get a progeny array composed of 4,875 individuals. The ~70 founders and all the progeny were genotyped using GBS. We also re-sequenced 20/70 founder lines (the others will be re-sequenced soon). Because of the high error rate of the GBS data, especially problematic for calling heterozygous sites, we employed a phasing and imputation strategy to infer the expected genotypes by combining parentage and GBS information.

This file is to document the `R` package we developed to solve the problems step by step.

## Install and Usage

Install devtools first, and then use devtools to install imputeR from github.

```{r, eval=FALSE}
# install and load devtools
devtools::install_github("hadley/devtools")
library(devtools)
# install and load imputeR
install_github("yangjl/imputeR")
library(imputeR)
```

## How to find help.
Within "R" console, type `?impute_mom` or `help(impute_mom)` to find help information about the function.

```{r}
library(imputeR)
?impute_mom
```

# Infer mom's genotype from GBS data

If we got mom's WGS data, this step can be skipped.

We have observed mom and observed (selfed) kids.  We want to know $P(G|\theta)$, or the probability of  mom's genotype given observed data $\theta$. And according to [Bayes' theorem](https://en.wikipedia.org/wiki/Bayes%27_theorem), 

$P(G|\theta) \propto P(\theta|G) \times P(G)$,   

where $P(G)$ is the probability of the genotype according to the Hardy-Weinberg equilibrium estimated from the population.
This consists of observed genotypes ($G'$) of both mom and kids. So:

$P(G|\theta)\propto \left( \prod\limits_{i=1}^{k}{P(G'_k|G)} \right) \times P(G'_{mom}|G) \times P(G)$,  

where $P(G'_{mom}|G)$ is the probability of mom's observed genotype given a true genotype $G$ by considering error rates, i.e. GBS homozygote error = 0.02 and heterozygote error = 0.8.  
And $P(G'_{k}|G)$ is the probability of the $k$ th kid’s observed genotype given genotype $G$ by considering error rates and Mendelian segregation rate. The function `impute_mom` was implemented to compute mom's genotype probabilities.

```{r, eval=FALSE}
# set the path
obs_mom <- c(0, 0, 0)
obs_kids <- list(c(0, 0, 0), c(0, 0, 0), c(0, 0, 0), c(0, 0, 0), c(0, 0, 0), c(0, 0, 0),
c(0, 0, 0), c(0, 0, 0), c(0, 0, 0), c(0, 0, 0), c(0, 0, 0), c(0, 0, 0))

geno <- impute_mom(obs_mom, obs_kids, hom.error=0.02, het.error=0.8)
momgeno(geno, oddratio=0.5, returnall=TRUE)
#install.packages(c("devtools", "roxygen2", "testthat", "knitr"))

```


This function is to impute mom's genotype from a progeny array of k kids at a single locus.
inferred_mom=1 -> 00, 2->01, 3->11


----------

# Imputing Founder Genotypes  
$P(G|\theta) \propto P(\theta|G) \times P(G)$   
$$P(G|\theta)\propto \left( \prod\limits_{i=1}^{k}{P(G'_i|G)} \right) \times \left( \sum_{n=1}^{mom} P(G'_{mom}|G) \right) \times P(G)$$  

----------

This function is to impute mom's genotype by finding the maximum likelihood of $P(G|\theta)$ from a progeny array of $k$ kids at a single locus.
- Where $\theta$ denotes observed data. It consists of observed genotypes ($G'$) of both mom and kids.  
- $P(G)$ is the probability of a genotype given Hardy-Weinberg frequencies estimated from the observed data.  
- $P(G'_{mom}|G)$ is the error matrix estimated from the data, i.e. homozygote error = 0.02 and heterozygote error =0.6.  
- $P(G'_i|G)$ is the error matrix times Mendelian segregation rate.  


----------

# Phasing Founder Genotypes  
$P(H|\theta) \propto P(\theta|H) \times P(H)$   
$P(H|\theta) \propto \left( \prod\limits_{i=1}^{k}{P(H'_k|H)} \right) \times P(H)$  
$P(H|\theta) \propto \left( \prod\limits_{i=1}^{k}\prod\limits_{l=1}^{n}{P(G'_{i,l}|H)} \right) \times P(H)$

- Where $\theta$ denotes observed data.
- $P(H)$ is the probability of the haplotype for a given window size of $n$.
- $P(G'_{i,l}|H)$ is the probability of kid $i$ at locus $l$ for a given haplotype $H$.
- The prior $P(H)$ is that all possible haplotypes of a given window size are equally likely.

----------

# Imputing and Phasing Kids  
$P(H_k|\theta) \propto P(\theta|H_k) \times P(H_k)$   
$P(H_k|\theta) \propto \left( \prod\limits_{i=k}{P(H'_k|H_k)} \right) \times P(H_k)$  
$P(H_k|\theta) \propto \left( \prod\limits_{i=k}\prod\limits_{l=1}^{n}{P(G'_{i,l}|H_k)} \right) \times P(H_k)$

- Where $\theta$ denotes observed data.
- $P(H)$ is the probability of the haplotype for a given window size of $n$.
- $P(G'_{i,l}|H)$ is the probability of kid $i$ at locus $l$ for a given haplotype $H$.
- The prior $P(H)$ is that all possible haplotypes of a given window size are equally likely.

```{r, eval=FALSE}
phase <- read.csv("../data/sim_phasing_res.csv")

hist(phase$er, breaks=30, main="Simulation (N=100)",col="#faebd7", xlab="Phasing Error Rate")
abline(v=mean(phase$er), col="red", lwd=2)
abline(v=median(phase$er), col="darkblue", lwd=2)
```

----------

# Phasing Dad of outcrossing progeny array  
$P(H_d|\theta) \propto P(\theta|H_d) \times P(H_d)$   
$P(H_d|\theta) \propto \left( \prod\limits_{i=1}^{k}{P(H'_k|H_d, H_m)} \right) \times P(H_m) \times P(H_d)$  
$P(H|\theta) \propto \left( \prod\limits_{i=1}^{k}\prod\limits_{l=1}^{n}{P(G'_{i,l}|H)} \right) \times P(H)$

- Where $\theta$ denotes observed data.
- $P(H_d)$ is the probability of the dad's haplotype for a given window size of $n$.
- $P(G'_{i,l}|H)$ is the probability of kid $i$ at locus $l$ for a given haplotype $H$.
- The prior $P(H)$ is that all possible haplotypes of a given window size are equally likely.


----------
